---
title: A random walk
author: Bernardo Reckziegel
date: '2018-07-27'
slug: a-random-walk
categories: []
tags:
  - time series
  - finance
  - simulation
---

This is my first post. I had planned to add some comments on how I created this blog using the `blogdown` package, but since I was revisiting a material related to time series I decided that writing about random walks would be a nice start. 

*** 

### 1. What is a Random Walk?

A random walk could be formally defined as a collection random series that are ordered in time. In a random walk (RW) the realized values of a time series are particular of that specific moment and cannot be extrapolated into the future. Needless to say, this imposes a lot of constraints to most traditional econometric models. 

One immediate implication is that if a particular series follows a random walk, no additional information can be learned from the data besides that we already have. This idea rests on the heart of the Efficient Market Hypothesis (EMH). 

A general description of a random walk process could be given by

`$$ y_t = \mu + y_{t - 1} + \epsilon_t $$`

The key insight is to manipulate this equation using the subscript `$t$` recursively. For example, in `$t - 1$` the above equation is

`$$ y_{t - 1} = \mu + y_{t - 2} + \epsilon_{t - 1} $$`

For `$t - 2$`

`$$ y_{t - 2} = \mu + y_{t - 3} + \epsilon_{t - 2} $$`

For `$t - 3$`

`$$ y_{t - 3} = \mu + y_{t - 4} + \epsilon_{t - 3} $$`

... and so on. 

Doing some algebra and substituting the second equation into the first we get

`$$ y_t = \mu + \underbrace{\mu + y_{t - 2} + \epsilon_{t - 1}}_{\text{Second Equation}} + \epsilon_t $$`

`$$ y_t = 2 \mu + y_{t - 2} + \epsilon_t + \epsilon_{t - 1} $$`

Doing the same thing for `$y_{t - 2}$` and `$y_{t - 3}$` (that is, putting `$y_{t - 2}$` inside `$y_{t - 1}$` and `$y_{t - 3}$` inside `$y_{t - 2}$`) we finally reach the result

`$$ y_t = 4 \mu +y_{t - 4} + \epsilon_t + \epsilon_{t - 1} + \epsilon_{t - 2} + \epsilon_{t - 3} $$`
    
Hopefully you can clearly see a pattern. The `$\mu$`'s increase with `$t$` and the impact of `$\epsilon_t$` accumulates indefinitely from it's starting point: *infinit memory*!

All the equations above can be divided in two different parts: deterministic and stochastic.

`$$ y_t =  \underbrace{\mu t + y_{t0}}_{\text{Deterministic}} + \underbrace{\sum{\epsilon_t}}_{\text{Stochastic}} $$`

The normal distribution is not a requirement for `$\epsilon$`, but is often used as a natural outcome of the Central Limit Theorem (CLT). 

From the empirical point of view this model imposes a challenge for those that aim to consistently beat the market: the value of `$y_{t}$` at `$y_{t + 1}$` is simply `$y_{t}$`. 


### 2. Create a function that generates a single random walk

The only parameters in the arithmetic random walk that need to be estimated are `$\mu$` and `$\sigma$`. The function bellow takes care of that thought the `mu` and `sd` arguments.


```{r message=FALSE, warning=FALSE, error=FALSE}

suppressPackageStartupMessages(library(tidyverse))

arithmetic_random_walk <- function(
    
    # function arguments
    initial_price = 100, 
    n             = 1000, 
    mu            = 0.0005, 
    sd            = 0.012, 
    seed          = FALSE

) {
    
    if (seed) set.seed(seed)
    
    random_walk <- 
        (initial_price + (mu * 100 * 1:n)) +    # deterministic
        cumsum(rnorm(n = n, sd = 0.0127 * 100)) # stochastic 
    
    random_walk 
    
}

```


The value of `mu` and `sd` were carefully chosen to emulate the long-term returns and volatility of S&P 500, around 12% and 20%, respectively.  

Fair enough. Now we have a formula to generate one process. The next step is generalize this procedure to as many replicates we want. 


### 3. Many paths

First a digression: one of the most important features of the `R` Programming Language is that each object that you create is a function. That means that, if you can use the plus sign (a function) to do `2 + 2`, you can also use your custom functions inside other functions as well. This is the key to leverage your skills in this language. 

In the example bellow, the function `multivariate_arithmetic_random_walk()` contains the same arguments of `arithmetic_random_walk()` but it's not constrained to execute only one path. The number of simulation can be controlled by the `n_simulations` argument. 

```{r}

many_arithmetic_random_walk <- function(n_simulations = 10, ...) {
    
    # tidy eval 
    dots_expr <- dplyr::quos(...)
    
    n_simulations %>%
        rerun(arithmetic_random_walk(!!! dots_expr)) %>% 
        as_tibble(validate = FALSE) %>% 
        `colnames<-`(., str_c('rw_', 1:n_simulations)) %>% 
        mutate(index = 1:nrow(.)) %>%
        select(index, everything()) %>% 
        gather(key, value, -index) %>% 
        mutate_if(is.character, as_factor)
    
}

```


Note how the `quos()` function (from the `dplyr` package) its used to do the tidy-eval *black-magic*: any arguments that you may want to change in `arithmetic_random_walk()` is passed to `...` and lazily evaluated through `!!!` (the *bang-bang-bang* operator). 


***

If you are curious about tidy-eval framework check this [webinar](https://www.rstudio.com/resources/webinars/tidy-eval/).

***


An interesting function that was used is `rerun()` from `purrr`. Basically, this is a functional that runs a script as many times you want. It is a easy way to get your code vectorized and to avoid the complexity of `for` loops.

`mutate_if()` transforms the character variables in factors, so them can be easily used in conjunction with `ggplot2`. The output contains all the simulated paths in a nice tidy `tibble` format. 


### 4. Plot the results

Now let's plot the results. The first step is to generate the paths. In second place, let's use `group_by()` to calculate the average simulated price at each point in time. The last step is just to modify the `ggplot2` layers according to our needs. 

```{r}

# generate and calculate the mean price in each point in time
mean_path <- 
    many_arithmetic_random_walk(n_simulations = 50) %>% 
    group_by(index) %>%
    mutate(mean = mean(value)) %>% 
    ungroup()

# plot
mean_path %>% 
    ggplot(aes(x = index, y = value, fill = key, alpha = 0.75)) + 
    geom_line(show.legend = FALSE, color = 'grey') + 
    labs(
        title    = 'Arithmetic Random Walk', 
        y        = 'Prices', 
        x        = 'Time Index'
    ) + 
    scale_y_continuous(labels = scales::dollar) +
    geom_line(aes(y       = mean),
              show.legend = FALSE,
              color       = 'steelblue',
              size        = 1
    ) +
    theme_classic()

```

