---
title: "A random walk"
author: "Bernardo Reckziegel"
date: '2018-07-27'
slug: a-random-walk
categories: []
tags:
- time series
- finance
- simulation
---

This is my first post. I had planned to add some comments on how I created this site using the `blogdown` [package](https://cran.r-project.org/web/packages/blogdown/), but since I was revisiting a material related to time series I decided that writing about random walks would be a nice start. 

*** 

### 1. What is a Random Walk?

A random walk (RW) could be formally defined as a collection random series that are ordered in time. In a random walk the realized values of a time series are particular of that specific moment and cannot be extrapolated into the future. This imposes a huge constraint to econometric theory. 

One immediate implication is that, if a particular series follows a random walk, no additional information can be learned from the data besides the one we already have at hand. This idea rests at the heart of the [Efficient Market Hypothesis (EMH)](https://en.wikipedia.org/wiki/Efficient-market_hypothesis). 

A general description of a random walk process could be given by:

`$$ y_t = \mu + y_{t - 1} + \epsilon_t $$`

The key insight is to manipulate this equation using the subscript `$t$` recursively. For example, in `$t - 1$` the above equation is:

`$$ y_{t - 1} = \mu + y_{t - 2} + \epsilon_{t - 1} $$`

For `$t - 2$`:

`$$ y_{t - 2} = \mu + y_{t - 3} + \epsilon_{t - 2} $$`

For `$t - 3$`:

`$$ y_{t - 3} = \mu + y_{t - 4} + \epsilon_{t - 3} $$`

... and so on. 

Doing some algebra and substituting the second equation into the first we get

`$$ y_t = \mu + \underbrace{\mu + y_{t - 2} + \epsilon_{t - 1}}_{\text{Second Equation}} + \epsilon_t $$`

`$$ y_t = 2 \mu + y_{t - 2} + \epsilon_t + \epsilon_{t - 1} $$`

Repeating the same step for `$y_{t - 2}$` and `$y_{t - 3}$` (that is, putting `$y_{t - 2}$` inside `$y_{t - 1}$` and `$y_{t - 3}$` inside `$y_{t - 2}$`) we finally reach the result:

`$$ y_t = 4 \mu +y_{t - 4} + \epsilon_t + \epsilon_{t - 1} + \epsilon_{t - 2} + \epsilon_{t - 3} $$`
    
Hopefully you can clearly see a pattern. The `$\mu$`'s increase with `$t$` and the impact of `$\epsilon_t$` accumulates indefinitely from it's starting point. The random walk has an infinite memory!

The equations above can be divided in two different parts: deterministic and stochastic.

`$$ y_t =  \underbrace{\mu t + y_{t0}}_{\text{Deterministic}} + \underbrace{\sum{\epsilon_t}}_{\text{Stochastic}} $$`

The normal distribution is not a requirement for `$\epsilon$`, but is often used as a natural outcome of the [Central Limit Theorem (CLT)](https://en.wikipedia.org/wiki/Central_limit_theorem). 

From the empirical point of view this model imposes a challenge for those that aim to consistently beat the market: the expected value of `$y_{t}$` for `$y_{t + 1}$` is simply `$ \mu t + y_{t}$`. 


### 2. Create a function that generates a single random walk

In the arithmetic random walk, the only parameters that needs to be estimated are `$\mu$` and `$\sigma$`. The function bellow generates the RW process, given an initial price, a time-series length, `mu` and `sd`.


```{r message=FALSE, warning=FALSE, error=FALSE}
suppressPackageStartupMessages(library(tidyverse))

arithmetic_random_walk <- function(
    # function arguments
    initial_price = 100, 
    n             = 1000, 
    mu            = 0.0005, 
    sd            = 0.012
) {
    random_walk <- 
        (initial_price + (mu * 100 * 1:n))  + # deterministic
        cumsum(rnorm(n = n, sd = sd * 100))   # stochastic 
    random_walk 
}

```


The value of `mu` and `sd` were carefully chosen to emulate the long-term returns and volatility of S&P 500, around 12% and 20%, respectively.  

The next step is generalize this procedure to as many replicates as we want. 


### 3. Many paths

In the example bellow, the function `multivariate_arithmetic_random_walk()` contains the same arguments of `arithmetic_random_walk()`, but it's not constrained to execute only one path. The number of simulation can be controlled by the `n_simulations` argument, while `...` absorbs any arguments that you may want to change in `arithmetic_random_walk()`. 

```{r}

many_arithmetic_random_walk <- function(n_simulations = 10, ...) {
    
    # tidy eval 
    dots_expr <- dplyr::quos(...)
    
    n_simulations %>%
        rerun(arithmetic_random_walk(!!! dots_expr)) %>% 
        as_tibble(validate = FALSE) %>% 
        `colnames<-`(., str_c('rw_', 1:n_simulations)) %>% 
        mutate(index = 1:nrow(.)) %>%
        select(index, everything()) %>% 
        pivot_longer(cols = -index) %>% 
        mutate_if(is.character, as_factor)
    
}

```

An interesting function that was used is `rerun()`, from `purrr`. It worth exploring if you wish to get your code vectorized and to avoid the complexity of `for` loops.


### 4. Plot the results

Now let's plot the results. First, I generated a wide diversity of paths that were stored in `mean_path` object. Them, I used `group_by()` to calculate the average simulated price at each point in time. Finally, added some custom layers to `ggplot2`. 

```{r}

# generate and calculate the mean price in each point in time
mean_path <- 
    many_arithmetic_random_walk(n_simulations = 50) %>% 
    group_by(index) %>%
    mutate(mean = mean(value)) %>% 
    ungroup()

# plot
mean_path %>% 
    ggplot(aes(x = index, y = value, fill = name, alpha = 0.75)) + 
    geom_line(show.legend = FALSE, color = 'grey') + 
    labs(
        title    = 'Arithmetic Random Walk', 
        y        = 'Prices', 
        x        = 'Time Index'
    ) + 
    scale_y_continuous(labels = scales::dollar) +
    geom_line(aes(y       = mean),
              show.legend = FALSE,
              color       = 'steelblue',
              size        = 1
    ) +
    theme_classic()

```

