---
title: Portfolio Analytics in the Tidyverse
author: Bernardo Reckziegel
date: '2019-02-20'
slug: portfolio-analytics-in-the-tidyverse
categories:
  - finance
  - R
tags:
  - optimization
  - estimation
  - tidyverse
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(PortfolioAnalytics)
library(tidyverse)
library(purrrlyr)
library(rsample)
library(timetk)


```


`Portfolio Analytics` (PA) is a well known library for portfolio optimization in the `R` community. Written before the `tidyverse` gain pulse, the PA package is often used in the "Base R" context. The objective of this post is to show that it can also be use inside the `tidyverse` (with all the power that comes with it!).

The basic usage of `portfolio Analytics` can be found [here](https://cran.r-project.org/web/packages/PortfolioAnalytics/index.html) and in it's four vignettes. I'll start direct from the tidy-world.

## Libraries

```{r, message=FALSE, warning=FALSE}
library(PortfolioAnalytics)
library(tidyverse)
library(purrrlyr)
library(rsample)
library(timetk)
library(tibbletime)
```

# Get the Data

`PortfolioAnalytics` comes the `edhec` dataset. Unfortunately, this data ends in 2009. To increase the number of observations lets download the full history of the EDHEC Hedge-Funds Index.

```{r}
edhec_updated <- 
    
    # read csv
    read_delim("https://risk.edhec.edu/sites/risk/files/indices/Indices/Edhec%20Alternative%20Indices/Web/table/history.csv", 
    delim = ";", 
    escape_double = FALSE, 
    col_types = cols(date = col_date(format = "%d/%m/%Y")), 
    trim_ws = TRUE
    ) %>% 
    
    # cleaning a beat
    dmap_if(.d = ., .p = is_character, .f = ~ str_remove(. , pattern = '%')) %>% 
    dmap_if(is_character, .f = ~ as.double(.) / 100) 

edhec_updated %>% 
    tail()
```

We now have gained ten additional years of data. 

## Tidy Portfolios

With a clean data at hand, let's `pipe` a simple mean-variance portfolio.

```{r}

# used latter in the optimization
edhec_updated_xts <- edhec_updated %>% 
    tk_xts(data = ., date_var = date, silent = TRUE) 

# You can pipe the portfolio specification!
mean_var_pspec <- portfolio.spec(assets = colnames(edhec_updated_xts)) %>%
  add.constraint(portfolio = ., type = "box", min = 0.00, max = 1.00) %>%
  add.objective(portfolio = ., type = "risk", name = "var") %>% 
  add.objective(portfolio = ., type = "return", name = "mean")
```

The first insight is that you can "pipe" the portfolios you want to work with. Second, the `tidyverse`offers a succinct way of making a reference to a _current objects_: the "`.`" (dot). The "dot" could also be ignored in this case, but I decided to keep it so the similarity with "Base R" could stand-out more clearly.

The optimization for the full period yields. 

```{r}
optim_port <- edhec_updated_xts %>% 
    optimize.portfolio(
        R               = ., 
        portfolio       = mean_var_pspec, 
        optimize_method = "ROI", 
        trace           = TRUE
)

optim_port


```

A monthly return of `r optim_port$objective_measures$mean * 100` and a standard-deviation of `r optim_port$objective_measures$StdDev * 100`.

## Rolling Tidy-Portfolios 

Suppose now that we would like to do this operation on a rolling-basis. It's possible to use `optimize.portfolio.rebalancing()`, but we would be outside the tidy-world. Instead, let's go tidy with full-power.

```{r}

opt_port_rolling <- edhec_updated %>% 
    rolling_origin(data        = ., 
                   initial     = 60, 
                   skip        = 0,
                   assess      = 1, 
                   cummulative = TRUE
)

opt_port_rolling

```

The object `edhec_updated` has `r nrow(edhec_updated)` rows while `opt_port_rolling` has only `r nrow(opt_port_rolling)`. That is, with the `rolling_origin` function from the `rsample` package we created a `tibble` with `r nrow(opt_port_rolling)` separated time-slices in which is possible to apply the optimization separably on a expanding window (`cumulative = TRUE`). The `rsample` [website](https://tidymodels.github.io/rsample/index.html) explain it's main functions in more detail. Worth while spending some time there! 

The important to remind thought, is that for every row we have now an `analysis()` (training) period and an `assessment()` (testing) period that can be directly assessed.


```{r}
opt_port_rolling$splits[[1]] %>% 
    analysis() %>% 
    tail()
```

This is the first period. 

In the second one we move forward by one month exactly as we would like to.

```{r}
opt_port_rolling$splits[[2]] %>% 
    analysis() %>% 
    tail()
```

And this process continues until the end of the series. 

## `purrr` them all!

`tidyverse` works best with `tibbles`. `PortfolioAnalytics`, on the other hand, requires an `xts` [object](https://cran.r-project.org/web/packages/xts/index.html). Let's do the required coercion (from `tibble` to `xts`) inside `map()` so we can continue working in the tidy-world. 

```{r}

start_purrr <- Sys.time()
map_opt_port_rolling <- opt_port_rolling %>%
    mutate(optimizations = map(
        .x = .$splits, 
        .f =  ~ analysis(.x) %>% 
            tk_xts(data = ., date_var = date, silent = TRUE) %>% 
            optimize.portfolio(
                R               = ., 
                portfolio       = mean_var_pspec, 
                optimize_method = "ROI", 
                trace           = TRUE
                )
            )
        )
end_purrr <- Sys.time()

map_opt_port_rolling

```

Every row of the `optimizations` column has an optimal portfolio for each period of time. 

```{r}

# for period 1
# using 60 realizations
map_opt_port_rolling$optimizations[[1]]

```

```{r}

# for period 2
# using 61 realizations
map_opt_port_rolling$optimizations[[2]]

```

```{r}

# for period 3
# using 62 realizations
map_opt_port_rolling$optimizations[[3]]

```

Notice that in period 2 the algorithm have not found a solution. There are ways to circumvent this problem inside the `tidyverse`. My preferred ones is by using `purrr`'s adverbs: `safely()`, `quietly()`, `possibly()` and `insistently()`. I'm not going to do that here, however, because the main objective is how to use the `PortfolioAnalytics` in the `tidyverse`, not to solve every possible bug. 

To extract the weights, returns and the standard-deviation let's continue mapping.

```{r}

map_opt_port_rolling %>% 
    mutate(weights = map(.x = optimizations, .f = ~ .x$weights), 
           std_dev  = map_dbl(.x = optimizations, .f = ~ .x$objective_measures$StdDev), 
           return  = map_dbl(.x = optimizations, .f = ~ .x$objective_measures$mean), 
           sharpe_ratio = return / std_dev
           ) 

```

Those are the _in-sample_ statistics. 

The true _out-of-sample_ statistics can be calculated by using the `assesment()` function.

```{r}
# true out-of_sample data for the 1 period optimization
opt_port_rolling$splits[[1]] %>% 
    assessment()
```

```{r}
# true out-of_sample data for the 2 period optimization
opt_port_rolling$splits[[2]] %>% 
    assessment()
```
As expected, every time we change the "splits" the data moves forward by one month. 

The `assessment` periods can be used to measure the amount of error in the estimation process, for example. Think of investors that are not be risk-averse, but _error-averse_ (see chapter 12 of [Robust Portfolio Optimization Management](https://www.amazon.com/Robust-Portfolio-Optimization-Management-Fabozzi/dp/047192122X/ref=sr_1_1?ie=UTF8&qid=1550689618&sr=8-1&keywords=robust+portfolio+optimization+and+management) for details).

## Who is Faster?

Lets address a final question: who is faster: `PortfolioAnalycis` or `purrr`? Let's find out.

```{r, message=FALSE, warning=FALSE}

pa_opt_time <- edhec_updated_xts %>% 
    optimize.portfolio.rebalancing(
                R               = ., 
                portfolio       = mean_var_pspec, 
                optimize_method = "ROI", 
                trace           = TRUE, 
                training_period = 60, 
                rolling_window  = NULL, 
                rebalance_on    = "months" ) %>% 
    .[["elapsed_time"]]

```

- `PortfolioAnalytics`: `r pa_opt_time` 
- `purrr` consumed `r end_purrr - start_purrr`

`purrr`is slower. For several of hundreds of assets in a granular periodicity this is definitely a drawback. It's interesting to investigate how [`furrr`](https://davisvaughan.github.io/furrr/) responds to tests like the one above. I still hold a preference for the "tidy-way" because of the great flexibility that gives for the user. 







